---
title: "K-means Cluster Analysis"
output: html_notebook
---

### Replication Requirements
To replicate this tutorial's analysis we need to load the following packages:
```{r}
library(tidyverse) # data manipulation
library(cluster) # clustering algorithms
library(factoextra) # clustering algorithms & visualization
```


### Data preparation
The build-in R dataset *USArrest* is used as demo data.
  - Remove missing data
  - Scale variables to make them comparable
```{r}
# load data
data('USArrests')
my_data <- USArrests

# Remove any missing value (i.e NA values for not available)
my_data <- na.omit(my_data)

# Scale variables
my_data <- scale(my_data)

# View the first 3 rows
head(my_data , n =3)
```

### Clarifying distance measures
The classification of obseravation into group, requires some methods measuring the distance or the (dis)similiarity between the observations. The result of this computation is known as a dissilimarity of distance matrix. There are many methods to calculate this distance information; the choice of distance measures is a critical step in clustering. It defines how the similarity of two elements (x,y) is calcuated and it will influence the shape of the clusters.

The choice of distance measures is a critical step in clustering. It defines how the similarity of two elements (x,y) is calculated it will influence the shape of the clusters. The classical methods for distance measures are *Euclidean* and *Manhattan distances*, which are defined as follow:
**Euclidean distance**:
      $d_euc(x,y) = \sqrt{\sum_{i=1}^{n}{(x_i-y_i)^2}}$ (1)
      
**Manhattan distance**:
      $d_man(x,y) = \sum_{i=1}^{n}{|x_i-y_i|}$
      
Where, x and y are two vectors of length n.

Other dissimilarity measures exist such as correlation-based distances, which is widely used for gene expression data analyses. Correlation-based distance is defined by substracting the correlation coefficient from 1. Different types of correlation methods can be used such as:

**Perason correlation distance**:
      $d_cor(x,y) = 1 - \frac{\sum_{i=1}^n{(x_i-\bar{x})(y_i-\bar{y})}}{\sqrt{\sum_{i=1}^n{(x_i-\bar{x})^2\sum_{i=1}^n(y_i-\bar{y})^2}}}$
      
**Spearman correlation distance**:
      $$
      
    
      
      
      
      
      
      
      
      
      
    