---
title: "ML_Finance"
output: html_notebook
---


### Introduction

Our goal is to investigate about stocks' behavior over a fixed period, i.e. 10 years range, from _2010-01-01_ to _2014-01-03_. We are runing *K-mean Cluster*, in distinguishing into difference group that base on each stock's fluctuation. Linear(or Log)-return is the easiest one that can illurstrate about the behavior of a stock over time. And it is easy to compute, which is the ratio between the price difference and the previous price.

We are using package *quantmod* , and function *getSymbol()* to download the data from the **yahoo** or **google finance**. So, we need to need all the Symbols of each stock, since different stocks have a unique abbrevation. 

```{r message=F}
# install.packages('quantmod')
## uncommond above to install quantomod
library(quantmod)
```

### Loading data

First, we need to load the symbols list:

```{r message=F}
NASDAQ <- read.delim("C:/Users/QIZHAOR/Desktop/PROGRAM/R/ML_Fin/NASDAQ.txt", header=FALSE)
symbols <-as.vector(factor(NASDAQ$V1))
```

Then, we need to define some function for computing the daily ruturn of the stock.
```{r}
daFa <- data.frame()
get.linRtn <- function(trick,From,To){
  symbol <- getSymbols(trick, from = From , to = To,
                       auto.assign = F,warnings = F)
  
  data <- as.data.frame(periodReturn(symbol, period = 'daily'))
  colnames(data) <- as.character(paste(trick,"Linrtn"))
  assign(trick,data, .GlobalEnv)
}
```

Since there are more than thousands of symbols and includind 10 years times, the data set is huge. One of the solution is to *Sampling* some stock for particularly investigating. We need to set up the sampling seed, for making our experiment be consistency.

```{r}
set.seed(400)
```

we are about to draw 30 symbols from the list at random.

```{r}
assets <- sample(symbols,30)
```

From now no, we are woring on these 30 stocks behaviors, and try to distinguish their behavior over a period of time.

We need to specify the time range:

```{r}
From <- '2018-01-01'
To <- '2019-01-01'
```

We are using for-loop and the function defined aboved to generate the data we want:

```{r}
for (asset in assets){
  get.linRtn(asset,From,To)
}
```



```{r}
my_function <- function(t){
  as.vector(data.frame(eval(parse(text = t))))
}

daFa <- as.data.frame(lapply(assets, my_function))
str(daFa)
```

























