---
title: "ML_loadData"
output: html_notebook
---

## Clustering in *S&P500*

Finally, we have nail down our taget is **S&P 500**. Our goal is to investigate the relationship between those stocks as a whole. The main theme of this project is working on clustering. Clustering of stocks base on their histortical data.


### Readying with data

We are using *quantmod* for downloading each stock's data. *getSymbols()* is a useful function for downloading the stock's historical prices and other useful data base on stock's unique **symbol**. 

First, we need to know all of *S&P500* stocks' symbols. We download a _CSV_ file which contains S&P 500 Companies with Financial Information. And we need to convert it into a _txt_ file, is easy to import into R. ( [Datahub](https://datahub.io/core/s-and-p-500-companies) ). 
```{r}
S.P500 <- read.delim("C:/Users/QIZHAOR/Desktop/PROGRAM/R/ML_Fin/S&P500.txt")
```

Take a look at the data, using *head()* function :

```{r}
head(S.P500)
```

We symbols neme from the *Symbol* column:

```{r}
Symbols <- sapply(S.P500$Symbol,as.character,simplify = 'array')

```

let's prepare the library we are using:
```{r message = FALSE}
library(quantmod)
```

Before going to download all 500+ data sets, we need to speicify some *date*, i.e. the starting date and the endding date. We need to specify, *From*, a character that stands for the starting date; and *To*, a character that stands for the endding date. 

We want to set, might be all of them start even later that, '1930-01-01' as the starting date; and set the '2019-08-01' as the endding month.

```{r}
From <- '1930-01-01'
To <- '2019-08-01'
```

Then, we are ready to load the data:

```{r}
Data <- sapply(Symbols,getSymbols,from = From , to = To, USE.NAMES = FALSE,simplify = 'array')
```

### Dicussion : Loading Error

In this area, we are going to arise issues that we have encounter with.
  - when are going to downloading stock's data by using their symbols, we found the listed below *symbols* are invaild:
    - "APC"
```{r}
## getSymbols("APC")
```
    So, one solution that is do not bother it :
```{r}
Symbols <- Symbols[Symbols != "APC"]
```
    - Another one *"BRK.B"*:
```{r}
#getSymbols("BRK.B")
```
```{r}
Symbols <- Symbols[Symbols != 'BRK.B']
```
    - *"BF.B"*
```{r}
Symbols <- Symbols[Symbols != 'BF.B']
```
    - *“DWDP”*
```{r}
Symbols <-Symbols[Symbols != "DWDP"]
```
    - *"DPS"*
```{r}
Symbols <- Symbols[Symbols != "DPS"]
```
    -*"GGP"*
```{r}
Symbols <- Symbols[Symbols != "GGP"]
```
    
    

    

    
    















